%% sections/baselines.tex
%% GOAL: a "reference baselines" section showing the benchmark is usable and discriminative.
%%   - non-LLM panel (clean-ceiling 0.83 / rule-based 0.00 / metric-heuristic judge 0.99):
%;        python3 bench/run_bench.py --baselines   (pure stdlib, no API)
%%   - a 4-model COMPREHEND capability ladder via DeepInfra (openai_compatible), k=8, all 7
%;        families, dev split, <$25: Llama-3.1-8B / Mistral-Small-24B / Llama-3.3-70B / DeepSeek-V4-Flash.
%% Gradient overall EM 0.55/0.74/0.77/0.84 (discriminates across the capability range) + per-family
%% structure: easy families saturate by the mid tier; agg_stats challenges even the strongest tested model
%% (DeepSeek 0.33) = real headroom; refactor forms its own semantic_ok ladder 0.73/0.83/0.95/0.99
%% (an EQUIVALENCE GATE, not optimal-form recovery); equivalence saturates only AT the strongest tested model,
%% and OOD scale still lowers it (~0.60). NOT a refactoring-quality/optimal-recovery saturation claim.
%% NUMBERS ARE REAL but recovered/validated: the live-run finalize hit two harness bugs (a java_gen
%% source-regen KeyError on the allowlist family, and a resume-aggregate bug giving a wrong _all),
%% so scores are re-graded from the persisted raw_outputs via
%; bench/ladder_analysis.py (dev ladder) / bench/g3_analysis.py (G3 tables) -> bench/out/ladder_comprehend.json, validated == the harness grader
%% on the clean DeepSeek batch (0.8385 == 0.8385). NOT the deferred LLM-judge study.

\section{Reference Baselines}
\label{sec:baselines}

We report two kinds of reference scores: a \emph{non-learned} panel that brackets the
task's floor and ceiling, and a small \emph{capability ladder} of open models that shows
the dataset discriminates across the capability range. These are reference points for a
resource, not a model evaluation. All numbers are on the public \texttt{dev} split (100
instances, 7 families, 5 languages, $k{=}8$ samples). The non-learned panel is produced by
\texttt{run\_bench.py -{}-baselines} (pure standard library, zero API); the model ladder is
four open models served through one OpenAI-compatible endpoint, graded by exact match on
the output-prediction (comprehension) task in a single cost-capped run (under \$25 of
API).

\begin{table}[t]
\centering
\caption{Non-learned reference panel on the public \texttt{dev} split, bracketing each
task. \textbf{Refactor} is semantic-gated simplification-quality (fraction of the
spaghetti$\rightarrow$clean gap closed, capped at the clean floor) on the rigorous Python
AST/MI lane ($n{=}300$ Python items); the language-agnostic \texttt{uniform\_lane} (lizard
cyclomatic complexity + token size + Buse--Weimer surface density) gives $0.95$ across all
five languages for the clean-ceiling. \textbf{Judge} is position-swap--controlled pairwise
accuracy over $n{=}250$ items. \textbf{Comprehend} is output-prediction exact match.
``---'' marks a (task, baseline) pair with no defined baseline.}
\label{tab:baselines}
\begin{tabular}{@{}lccc@{}}
\toprule
Baseline & Refactor & Judge (pairwise acc.) & Comprehend (EM) \\
\midrule
Clean-ceiling          & $0.83$ & ---    & ---    \\
Rule-based             & $0.00$ & ---    & ---    \\
Metric-heuristic judge & ---    & $0.99$ & ---    \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\caption{Capability ladder on the comprehension (output-prediction) task --- exact match,
$k{=}8$, over the public \texttt{dev} split, for four open models served through one
OpenAI-compatible endpoint. Overall exact match rises monotonically with model capability,
so the dataset \emph{discriminates across the capability range}. The per-family breakdown
shows the structure: ``easy'' families saturate by the mid tier, while \texttt{agg\_stats}
--- predicting the result of an \texttt{AGGREGATE} (sum/min/max) over a large integer
collection --- challenges even the strongest model we tested, leaving genuine headroom. Outputs are
graded by exact match against the oracle. (Two harness finalize/resume bugs in the
live-run path were caught during this run; the numbers are re-graded from the persisted
raw completions, validated to match the harness grader on a clean batch.)}
\label{tab:ladder}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Comprehend EM & {\footnotesize Llama-3.1-8B} & {\footnotesize Mistral-24B}
              & {\footnotesize Llama-3.3-70B} & {\footnotesize DeepSeek-V4-Flash} \\
\midrule
\textbf{Overall}              & $0.55$ & $0.74$ & $0.77$ & $0.84$ \\
\midrule
\texttt{agg\_stats} (hard)    & $0.00$ & $0.10$ & $0.09$ & $0.33$ \\
\texttt{threshold\_select}    & $0.42$ & $0.60$ & $0.79$ & $1.00$ \\
\texttt{fsm\_transition}      & $0.05$ & $0.61$ & $0.66$ & $0.54$ \\
\texttt{config\_resolver}     & $0.71$ & $0.99$ & $1.00$ & $1.00$ \\
\texttt{status\_router} (easy)& $1.00$ & $1.00$ & $1.00$ & $1.00$ \\
\bottomrule
\end{tabular}
\end{table}

The non-learned panel brackets the refactor task. The \emph{clean-ceiling} --- the
engine's own idiomatic clean rendering, graded through the same semantic gate and
metric as a model --- reaches $0.83$ AST simplification-quality (Python) and $0.946$ on
the language-agnostic uniform lane across all five languages, with a $1.00$
semantic-equivalence rate, and lands inside the optimality band on every item, confirming
the labelled optimum is reachable in every backend and the metric is not saturated at $1.0$. At the other
end, two non-learned references both score $0.00$: a pure autoformatter (lower bound;
semantics-preserving but removes no anti-patterns) and a conservative AST
\emph{rule-based} simplifier (constant-folding and literal dead-branch removal). Both stay
semantically correct yet recover essentially none of the construction gap, so the task is
non-trivial and demands genuine restructuring rather than reformatting. For the judge
task, the \emph{metric-heuristic judge} --- picking the lower static-complexity candidate
of each pair --- attains $0.99$ position-consistent pairwise accuracy (Spearman
rank-vs-complexity $\rho \approx -0.99$). This is a deliberately strong non-LLM reference:
because complexity is monotone with the construction knob \emph{by design}, any useful LLM
judge must be measured \emph{against} this floor, not merely above chance.

\paragraph{An open-model capability ladder.}
The four-model ladder (\Cref{tab:ladder}) shows the dataset \emph{discriminates}. On
the comprehension task, overall exact match rises monotonically with model capability ---
$0.55$ (Llama-3.1-8B) $\to 0.74$ (Mistral-24B) $\to 0.77$ (Llama-3.3-70B) $\to 0.84$
(DeepSeek-V4-Flash) --- so the benchmark separates models across the capability range, not
merely above a trivial floor. The ladder is also language-invariant for the capable models:
per-language comprehension exact match varies by only $\le 0.04$ across all five languages
for the top three models, confirming the five-language corpus is genuinely parallel; only
the weakest model is language-sensitive (a $0.22$ spread). The per-family structure is the
more informative finding.
``Easy'' families (e.g.\ \texttt{status\_router}, \texttt{config\_resolver}) are solved by
the mid tier and \emph{saturate}, contributing discrimination only at the low end --- a
ceiling effect we report explicitly. Others do not saturate:
\texttt{threshold\_select} grades monotonically across the range, \texttt{fsm\_transition}
discriminates but not monotonically (the strongest model does not dominate every family),
and \texttt{agg\_stats} --- which asks a model to predict the result of an
\texttt{AGGREGATE} over a collection of up to $160$ integers --- challenges
\emph{every} model, leaving the strongest model we tested, DeepSeek-V4-Flash, at only $0.33$. That a
controlled, regenerable, contamination-resistant family stumps the strongest model we tested is exactly
the headroom a forward-looking resource needs, and it mirrors how controlled generation has
exposed arithmetic-reasoning limits elsewhere~\citep{mirzadeh2024gsm}. The
contamination resistance is now \emph{measured}, not merely designed: re-minting the
\texttt{dev} families into a private held-out \texttt{test} split with fresh secret values
leaves the ladder essentially unchanged --- per-model comprehension exact match on the
Tier-A re-mint tracks the \texttt{dev} split to within $|\Delta|\le 0.012$, and the refactor
semantic-equivalence ladder matches \texttt{dev} across all four models ($|\Delta|\le 0.011$;
strongest model $\Delta=-0.0001$) --- even though every prompt hash differs.
The tasks differ in \emph{what kind} of competence they probe.
\emph{Refactoring} forms its own capability ladder in semantic-equivalence rate (the
fraction of $k{=}8$ rewrites that compile, run, and match the oracle), rising $0.73 \to 0.83
\to 0.95 \to 0.99$ across the same four models, so it discriminates at the lower tiers
while the strongest model's restructurings are almost always behavior-preserving. This
rate is an equivalence \emph{gate} (it credits any semantics-preserving rewrite, including
near-verbatim echoes), not a measure of optimal-form recovery, so we do \emph{not} claim a
model recovers the labelled optimum; equivalence is also not uniform with difficulty ---
far out-of-distribution scale and depth (beyond the in-range $W\le160$ sweep) lower even
the strongest model's equivalence rate to $\approx 0.60$. With refactor equivalence otherwise near-saturated for the strongest models, the
\emph{additional} comprehension headroom there is concentrated in the arithmetic
\texttt{agg\_stats} family, with the judge meta-evaluation as future work. The resource's contribution is the controlled, by-construction-labelled
data; model ranking is one downstream use, and the ladder above is a reference point for
it, not this paper's claim.

\paragraph{The intrinsic scale knob breaks even the strongest model we tested.}
Because difficulty is \emph{tunable} along the intrinsic axis, we can locate exactly where a
model fails. The \texttt{agg\_stats} family asks a model to predict the result of an
\texttt{AGGREGATE} (sum/min/max) over a collection of $W$ integers; sweeping $W$ from $8$ to
$160$ at fixed semantics is a controlled difficulty gradient. Exact match collapses with $W$
for \emph{every} model, the strongest included (\Cref{tab:scaling}): DeepSeek-V4-Flash is perfect
at $W{=}8$ but reaches near-zero by $W{\approx}48$, and all four models are at $0$ by
$W{=}64$ --- they cannot reliably aggregate more than a few dozen integers. The contrast with
the lookup family \texttt{config\_resolver} is sharp: there, every mid-tier-and-up model
stays at $\approx 1.00$ across all $N$ (no cliff), confirming the \texttt{agg\_stats} collapse
is arithmetic-reasoning difficulty driven by problem \emph{size}, not a generic artifact. A
second, sharper control comes from running the \emph{refactor} task on the \emph{same}
\texttt{agg\_stats} programs: there the semantic-equivalence rate is flat and high across
the entire $W$ sweep ($\approx 1.00$ for the capable models from $W{=}8$ to $W{=}160$),
even where comprehension exact match has fallen to $0$. Restructuring a wide aggregation is
width-invariant; \emph{evaluating} the aggregate is width-sensitive --- so on identical
inputs the resource cleanly separates a structural competence from a computational one. This
is the intrinsic-difficulty axis doing its job. The \emph{incidental} axis behaves oppositely:
comprehension exact match is flat across the messiness knob (minimal/standard/max) for every
model, so on this task the difficulty is \emph{intrinsic} (problem size), not
presentational, as the orthogonal-axes design intends. And it gives the resource
forward-looking headroom: a contamination-controlled, regenerable arithmetic-aggregation probe
that capable open models fail past a small size --- the controlled-generation analogue of
how \textsc{GSM-Symbolic} exposed brittle math reasoning~\citep{mirzadeh2024gsm}.

\begin{table}[t]
\centering
\caption{Comprehension exact match on \texttt{agg\_stats} as a function of the collection
size $W$ (the intrinsic scale knob), $k{=}8$, \texttt{dev} split. Difficulty is tunable and
breaks every model with scale: the strongest model we tested, DeepSeek-V4-Flash is perfect at $W{=}8$ and at $0$
by $W{=}64$. The lookup family \texttt{config\_resolver} shows no such cliff (all
mid-tier-and-up models stay at $\approx 1.00$ across $N$), so the collapse is reasoning
difficulty, not a generic effect. Per-cell $n\approx20$ items; the full $W$ curve and the
other families are in the released \texttt{ladder\_scaling.json}.}
\label{tab:scaling}
\begin{tabular}{@{}lrrrrr@{}}
\toprule
\texttt{agg\_stats} EM & $W{=}8$ & $W{=}16$ & $W{=}32$ & $W{=}64$ & $W{=}128$ \\
\midrule
Llama-3.1-8B      & $0.00$ & $0.00$ & $0.00$ & $0.00$ & $0.00$ \\
Mistral-24B       & $1.00$ & $0.17$ & $0.01$ & $0.00$ & $0.00$ \\
Llama-3.3-70B     & $0.72$ & $0.04$ & $0.14$ & $0.00$ & $0.00$ \\
DeepSeek-V4-Flash & $1.00$ & $0.69$ & $0.17$ & $0.00$ & $0.00$ \\
\bottomrule
\end{tabular}
\end{table}
