%% sections/intro.tex
%% Data-centric (DMLR) framing: foreground value to ML research, not just SE tooling.

\section{Introduction}
\label{sec:intro}

Source code is one of the most heavily used data modalities in machine learning, and
the dominant way to obtain it is to mine it: scrape repositories, filter, deduplicate,
and ship. Mined corpora are abundant, but they are \emph{uncontrolled}. A snippet's
semantics, its surface ``messiness,'' and its difficulty are whatever the wild happened
to contain; there is no known-optimal reference to grade an instance against; and the
two things a study most often wants to vary --- \emph{how hard the problem is} and
\emph{how poorly it is written} --- are confounded, because in real code a longer
program is usually also a messier one. Worse, every public sample is a moving target for
contamination: any snippet a researcher uses to evaluate a code model may already sit in
that model's training set, so a high score can reflect memorization rather than
capability~\citep{jacovi2023stop,riddell2024quantifying}. Mining gives us volume; it does not give us
\emph{control}, \emph{labels}, or \emph{freshness}.

A growing set of data-centric ML problems need exactly that control. Contamination-aware
evaluation of code models requires items that can be \emph{re-minted} after a training
cutoff rather than drawn from a static, leakable pool. Training and probing studies want
parallel data in which one factor moves while the rest are held fixed. Robustness
analysis wants to ask whether a model degrades because \emph{the problem got bigger} or
because \emph{the presentation got messier} --- a question no fixed corpus can answer,
because the two grow together in the wild. Refactoring, clone-detection, and
code-translation research wants paired clean/messy programs whose semantic equivalence is
\emph{guaranteed}, not merely assumed from a commit message. Each of these is a request
for a piece of data that mined corpora structurally cannot supply: a labelled,
equivalence-controlled, freshly-mintable, multi-language instance.

We present \tool, a tool that synthesizes such data. \tool{} is an
\emph{anti-optimization transpiler}: where an optimizing compiler holds semantics fixed
and \emph{minimizes} a cost, \tool{} holds semantics fixed and \emph{maximizes} incidental complexity. It maps a clean, language-agnostic JSON
intermediate representation (IR) to deliberately redundant, fully-flattened programs in
five languages (Python, JavaScript, Go, Java, C++), and every emitted program is
compiled, run, and checked against a reference oracle, so each instance is correct
\emph{by construction}. Because the clean IR is a known-optimal reference and
``messiness'' is dialed by composable, strictly-nested anti-pattern profiles, each
instance carries labels along two \emph{orthogonal} difficulty axes --- \emph{intrinsic}
(the problem grows) and \emph{incidental} (presentation degrades at fixed
semantics) --- and a single IR yields a parallel corpus across all five languages. To
resist contamination, scored instances are minted fresh from a \emph{private} held-out
seed that is never committed, a public canary GUID enables training-set audits, and
regenerable held-out tiers turn ``is this score inflated by memorization?'' into a measured
quantity (by comparing public-split scores against freshly-minted private-split scores)
rather than a hope. A meta-evaluation of whether LLM judges track a code-quality order known independently of
any rater is \emph{one} application this resource enables, and we use it to motivate the
design; this paper delivers and documents the \emph{instrument} and the dataset it
produces, which stand on their own as a contribution. Because the deliverable is a
\emph{generator}, the resource's value is its regenerability, not the size of any one
frozen download: the public \texttt{dev} split is a compact, fully-inspectable reference
($100$ instances), while an unbounded stream of fresh, equivalently-labelled instances
(including the private scored split) is available on demand from the same
contract. We therefore treat instance count as a tunable output, not a headline number.

\paragraph{Contributions.} This paper contributes:
\begin{itemize}
  \item \textbf{The generator and its correctness-by-construction design}
        (\Cref{sec:resource}): a staged IR$\rightarrow$five-language pipeline in which
        valid syntax is structural and runtime safety is always-on, gated by an oracle
        validator so that every released instance provably builds and computes the
        intended result.
  \item \textbf{Two orthogonal, by-construction difficulty labels}
        (\Cref{sec:axes}): an \emph{incidental} messiness knob (a strictly-nested
        anti-pattern ladder) crossed with an \emph{intrinsic} problem-size knob, so a
        score change can be attributed to one cause with the other held fixed --- a
        decomposition no mined corpus offers.
  \item \textbf{A contamination protocol} (\Cref{sec:contam}): mint-after-cutoff
        generation, a public \texttt{dev} split versus a private held-out \texttt{test}
        split, an embedded canary, and graded A/B/C held-out tiers (regenerable held-out
        structures for contamination-resistance) --- with the public/private asymmetry as
        the mitigation, not an apology.
  \item \textbf{Construct-validity evidence} (\Cref{sec:validity}) that the
        by-construction quality order moves established complexity and readability
        metrics, reported with its weaknesses stated plainly.
  \item \textbf{Reference baselines} (\Cref{sec:baselines}) on a four-model open ladder
        that establish the dataset is usable and discriminative: comprehension and
        refactoring each trace a capability ladder across model scale; public-\texttt{dev}
        scores match freshly-minted private-\texttt{test} scores, turning the contamination
        protocol from a design claim into measured evidence; and on identical programs,
        refactoring equivalence is invariant to intrinsic problem size for the capable models
        while comprehension's
        arithmetic aggregation collapses with it even for the strongest model we tested, so the dataset
        separates a structural from a computational competence. Released as a reusable, FAIR,
        dependency-free open-source artifact (to be archived under a DOI at release).
\end{itemize}

\paragraph{Artifact.} \tool{} is open source under the MIT license, depends only on the
Python standard library, ships runnable examples, a test suite, and a documented
extension contract, and will be archived under a DOI; a ``Datasheet for Datasets''
(\Cref{app:datasheet}) accompanies the dataset.
