%% sections/related.tex
%% Data-centric (DMLR) related work. Land the four differentiators.

\section{Related Work}
\label{sec:related}

We position \tool{} as a \emph{data resource} relative to five neighboring lines of work:
data-documentation norms; code/LLM benchmarks and the contamination problem;
contamination-\emph{detection} methods; procedurally-generated and diagnostic benchmarks
(in code and other modalities); and the controlled-transformation literature (obfuscation,
mutation, code smells, readability) from which the generator borrows its primitives.

\paragraph{Data documentation.} The data-centric ML community has converged on
structured documentation of provenance, composition, and intended use: Datasheets for
Datasets~\citep{gebru2021datasheets}, Model Cards~\citep{mitchell2019model}, and Data
Cards~\citep{pushkarna2022datacards} ask a dataset's (or model's) authors to make its
construction and limitations legible. We follow that norm rather than extend it: the dataset ships with a
datasheet (\Cref{app:datasheet}). What is unusual here is that, because the data is
\emph{synthesized} from a documented generator rather than collected, the answers to
many datasheet questions (how instances were labelled, whether they were preprocessed,
whether they can be regenerated) are exact and machine-checkable rather than
best-effort recollections.

\paragraph{Code and LLM benchmarks, and contamination.} Functional-correctness benchmarks
--- \textsc{HumanEval}~\citep{chen2021evaluating}, \textsc{MBPP}~\citep{austin2021program},
\textsc{APPS}~\citep{hendrycks2021apps}, the multi-task
\textsc{CodeXGLUE}~\citep{lu2021codexglue}, and rigorous re-tests that catch silently-wrong
solutions~\citep{liu2023evalplus} --- grade \emph{what} a model produces but fix
\emph{presentation} and, being static and public, are exposed to training-set
contamination~\citep{sainz2023nlp,riddell2024quantifying,jacovi2023stop}: a model may
score well because it has seen the items. A parallel line develops \emph{detectors} for
this leakage --- membership-inference probes such as Min-K\%~\citep{shi2024detecting},
exchangeability tests with false-positive guarantees~\citep{oren2024proving}, and the
training-data-extraction and memorization analyses that motivate canaries in the first
place~\citep{carlini2021extracting} --- a-posteriori tools that complement, rather than
replace, a generator that can simply mint uncontaminated data on demand. Two structural
responses have emerged. One gates
on \emph{time}, scoring only items released after a model's training cutoff, as in the
continuously-updated \textsc{LiveCodeBench}~\citep{jain2024livecodebench} and
\textsc{LiveBench}~\citep{white2024livebench}; the other makes instances
\emph{regenerable} so a fresh, never-seen draw is always available. \tool{} adopts both
--- a private held-out split minted after the cutoff plus a public canary
(\Cref{sec:contam}) --- but differs from time-gated suites in a way that suits a data
\emph{resource}: where a living benchmark \emph{harvests} new real-world problems over
time (and so cannot supply a known-optimal label or hold semantics fixed along a
difficulty axis), \tool{} \emph{synthesizes} its scored data, so every fresh draw still
carries by-construction labels and the two orthogonal axes. The scored data is thus not a
fixed file to be protected but a fresh, fully-labelled sample to be minted --- the only
durable defense once a public split inevitably leaks.

\paragraph{Procedurally generated and diagnostic benchmarks.} Synthesizing evaluation data
from a controlled generator is well established \emph{outside} code, precisely because it
buys control that scraping cannot: bAbI generates reasoning tasks from
templates~\citep{weston2015babi}, CLEVR renders visual-reasoning questions from a scene
grammar~\citep{johnson2017clevr}, SCAN emits compositional command/action pairs from a
small grammar~\citep{lake2018scan}, and \textsc{GSM-Symbolic} instantiates grade-school
math problems from symbolic templates to expose memorization and brittle
variance~\citep{mirzadeh2024gsm}. The same controllability now drives \emph{training}-data
synthesis for code: textbook-style synthetic corpora~\citep{gunasekar2023textbooks} and
open-source-seeded instruction data~\citep{wei2024magicoder} trade raw scale for control.
\tool{} brings this procedural-generation stance to a \emph{multi-language,
oracle-verified, by-construction-labelled} code resource: like these generators it
controls the stimulus, but unlike them it additionally guarantees semantic equivalence by
execution and carries a known-optimal reference for every instance.

\paragraph{Controlled code transformation.} The moves \tool{} composes are not new in
isolation. Source obfuscation deliberately degrades readability while preserving
behavior; the eleven \texttt{SPAGH\_*} transforms descend from the Collberg--Thomborson
taxonomy of obfuscating transformations~\citep{collberg1997taxonomy} (opaque predicates,
redundant computation) and from classic code smells~\citep{fowler-refactoring}. Code
readability and complexity have their own measurement traditions --- from Halstead's
software-science measures~\citep{halstead1977elements} through the learned surface-feature
readability metric of Buse and Weimer~\citep{buse2010readability} and its textual-feature
successor~\citep{scalabrino2018comprehensive} --- which we re-use as external validity
witnesses (\Cref{sec:validity}) rather than as generators. The defining difference is one of
\emph{purpose and labelling}. An obfuscator emits degraded code but no quality
\emph{order} and no known-optimal target; a code-smell injector or a mutation
tool~\citep{demillo1978hints,jia2011analysis} perturbs a program but (in the mutation case)
deliberately \emph{breaks} semantics to test a suite --- the inverse of our
equivalence-preserving goal; a readability metric scores existing code but does not generate a
graded corpus. \tool{} reuses these primitives as a \emph{maintainability-degradation
knob with ground truth}: a strictly-nested ladder that is, by construction, a quality
order at fixed semantics.

\paragraph{The four differentiators.} Each neighbor supplies some of the affordances a
data-centric study needs; none supplies all four at once.
\begin{itemize}
  \item \emph{Oracle-verified semantic equivalence.} Every instance --- and, for the
        refactoring use case, every candidate rewrite --- is compiled, run, and checked
        against a reference oracle, so equivalence is established by execution rather than
        assumed from provenance. Obfuscators and smell injectors typically assert
        equivalence; mutation tools invert it.
  \item \emph{By-construction graded quality order.} The incidental knob is a
        strictly-nested anti-pattern inclusion chain, so the messy$\rightarrow$clean order
        is known \emph{independently of any rater}. Obfuscators give a binary
        clean/obfuscated contrast; readability metrics give a rater-dependent score, not a
        ground-truth order.
  \item \emph{Multi-language parallel corpus.} One IR emits the same logic in five
        languages, yielding aligned cross-language instances. Obfuscation and mutation
        tools are typically single-language.
  \item \emph{Contamination control.} Mint-after-cutoff generation with a private
        held-out split, a canary, and regenerable held-out structure tiers (\Cref{sec:contam}). Static
        benchmarks and the transformation tools carry no contamination protocol.
\end{itemize}
\noindent To our knowledge no prior generator combines oracle-verified equivalence, a
by-construction graded quality order, a multi-language parallel corpus, and a
contamination protocol in a single artifact; that combination is the resource's novelty.
