%% sections/resource.tex
%% GOAL: describe the RESOURCE — the generator pipeline, the operation/
%% anti-pattern model, the two orthogonal difficulty axes, the oracle/validation, and the
%% contamination protocol.

\section{The Resource}
\label{sec:resource}

\subsection{Pipeline overview}
\tool{} is a staged pipeline that turns a clean, language-agnostic JSON IR into an
oracle-validated, two-axis-labelled multi-language dataset (\Cref{fig:pipeline}).

\begin{figure}[t]
\centering
\resizebox{\linewidth}{!}{%  scale the picture to fit the (narrower) DMLR column width
\begin{tikzpicture}[
  font=\footnotesize,
  node distance=5mm and 6mm,
  every node/.style={align=center},
  stage/.style={draw, rounded corners, fill=blue!7, inner sep=3pt, minimum height=8.5mm, text width=14mm},
  store/.style={draw, cylinder, shape border rotate=90, aspect=0.16, fill=gray!12, inner sep=3pt, minimum height=9mm, text width=14mm},
  chip/.style={draw, rounded corners=1.5pt, fill=white, inner xsep=3pt, inner ysep=1.8pt, font=\scriptsize, minimum width=14mm},
  note/.style={font=\scriptsize, inner sep=1.5pt, text width=20mm},
  flow/.style={->, semithick},
]
% input
\node[store] (ir) {JSON IR};
\node[note, below=1pt of ir, text width=22mm] (ops) {known-optimal reference; 4 op primitives};
% parser
\node[stage, right=of ir] (parser) {Parser};
\node[note, below=1pt of parser, text width=17mm] (pnote) {validate IR $\to$ typed model};
% planner
\node[stage, right=8mm of parser] (planner) {Planner};
% generators
\node[chip, right=10mm of planner] (g3) {Go};
\node[chip, above=1.6pt of g3] (g2) {JavaScript};
\node[chip, above=1.6pt of g2] (g1) {Python};
\node[chip, below=1.6pt of g3] (g4) {Java};
\node[chip, below=1.6pt of g4] (g5) {C++};
\node[note, above=1.4pt of g1] (genlbl) {Generators ($\times5$)};
\node[draw, rounded corners, dashed, fill=none, inner sep=3pt, fit=(genlbl)(g5)] (gen) {};
% validator
\node[stage, right=8mm of gen] (val) {Validator\\+ Oracle};
\node[note, below=1pt of val, text width=23mm] (vnote) {compile, run, compare to oracle (missing $\to$ SKIP)};
% anti-pattern database / messiness knob feeding the planner
\node[store, below=11mm of planner, text width=36mm, font=\scriptsize] (db)
  {Anti-pattern database \& messiness knob:\\clean\,$\subset$\,minimal\,$\subset$\,light\\$\subset$\,standard\,$\subset$\,heavy\,$\subset$\,max\\\textit{(fixed semantics)}};
% output splits
\node[chip, right=9mm of val, text width=30mm] (dev) {\textbf{dev} split:\\public, committed};
\node[chip, below=2pt of dev, text width=30mm] (test) {\textbf{test} split:\\private, held-out};
\node[note, above=1.4pt of dev] (datalbl) {Labelled dataset};
% background fills (drawn behind node contents so labels stay visible)
\begin{scope}[on background layer]
\node[draw, rounded corners, fill=green!7, inner sep=3pt, fit=(datalbl)(dev)(test)] (data) {};
\node[draw, rounded corners, fill=gray!5, inner sep=6pt, fit=(parser)(pnote)(planner)(gen)(val)(vnote)(db)] (tool) {};
\end{scope}
% flow
\draw[flow] (ir) -- (parser);
\draw[flow] (parser) -- (planner);
\draw[flow] (planner) -- (gen);
\draw[flow] (gen) -- (val);
\draw[flow] (val) -- (data);
\draw[flow] (db) -- (planner);
% annotations
\node[note, text width=40mm] at ([yshift=2.3mm]tool.north) {\tool{} (anti-optimization transpiler)};
\node[note, below=1pt of data, text width=34mm] (axes) {2 orthogonal axes: \emph{intrinsic} size (N/L/W/T) $\times$ \emph{incidental} messiness};
\end{tikzpicture}%
}% end \resizebox
\caption{The \tool{} pipeline. A clean, language-agnostic JSON IR---built from four
operation primitives (\texttt{MEMBERSHIP\_CHECK}, \texttt{KEY\_VALUE\_LOOKUP},
\texttt{AGGREGATE}, \texttt{CONDITIONAL\_SELECT}) and serving as the known-optimal
reference---is validated by the \emph{Parser} into a typed model, so a malformed IR can
never reach a generator. The \emph{Planner} then selects an anti-pattern profile from a
data-driven database via the strictly-nested six-level messiness knob
(\textsf{clean} $\subset$ \textsf{minimal} $\subset$ \textsf{light} $\subset$ \textsf{standard} $\subset$ \textsf{heavy} $\subset$ \textsf{max})
\emph{at fixed semantics}; five \emph{Generators} emit the deliberately redundant,
fully-flattened programs (Python, JavaScript, Go, Java, C++); and the \emph{Validator}
compiles and runs each target against a reference \emph{oracle} (a missing toolchain
\textsf{SKIP}s, never a silent pass). The output is a dataset labelled along two
orthogonal difficulty axes---\emph{intrinsic} problem size ($N$/$L$/$W$/$T$) and
\emph{incidental} presentation (the messiness knob at fixed semantics)---split into a
public, committed \texttt{dev} split and a private \texttt{test} split minted from a
never-committed held-out seed.}
\label{fig:pipeline}
\end{figure}

The \emph{Parser} validates the IR into a typed model (so a malformed IR can never reach
a generator). The \emph{Planner} reads a data-driven anti-pattern database and selects
which transforms apply per operation at the chosen \emph{profile}. Each of five
\emph{Generators} emits source through a \texttt{CodeEmitter} that makes correct
indentation/bracing \emph{structural} rather than hand-managed, with always-on safety
(try/catch + guards + a pre-set fallback) injected via hooks. The \emph{Validator}
compiles and runs each target and compares its output to a reference \emph{oracle}.

\paragraph{Design properties.}
\begin{itemize}
  \item \emph{Correctness by construction.} Valid syntax is structural; safety is always
        on, so output is crash-free regardless of profile strength.
  \item \emph{Provable correctness.} Every target is compiled, run, and checked against
        the oracle; a missing toolchain \textsf{SKIP}s (never a silent pass).
  \item \emph{Deterministic.} The same IR yields byte-identical output (enabling
        golden-snapshot testing and reproducible datasets).
  \item \emph{Data-driven.} The anti-patterns and strength profiles live in a JSON
        database, not in code.
\end{itemize}

\subsection{Operations and anti-patterns}
\label{sec:ops}
The IR composes a sequence of typed \emph{operations} (Table~\ref{tab:ops}); each reads
validated inputs and writes a fresh result variable.

\begin{table}[t]
\centering
\caption{Supported operations (v0.2).}
\label{tab:ops}
\begin{tabular}{@{}lll@{}}
\toprule
Operation & Semantics & Result \\
\midrule
\texttt{MEMBERSHIP\_CHECK}   & \texttt{target in collection}              & bool \\
\texttt{KEY\_VALUE\_LOOKUP}  & \texttt{map.get(key, default)}             & scalar \\
\texttt{AGGREGATE}           & \texttt{sum|min|max(int collection)}       & int \\
\texttt{CONDITIONAL\_SELECT} & \texttt{then if (a$\,\bowtie\,$b) else else}& scalar \\
\bottomrule
\end{tabular}
\end{table}

The eleven \texttt{SPAGH\_*} transforms span structural, verbosity, control, and layout
categories; three are grounded in the Collberg--Thomborson obfuscation
taxonomy~\citep{collberg1997taxonomy} and classic code smells~\citep{fowler-refactoring}.
They compose into three strictly-nested profiles
(\textsf{minimal}~$\subset$~\textsf{standard}~$\subset$~\textsf{max}); the released
dataset uses a finer six-level knob
(\textsf{clean}~$\subset$~\textsf{minimal}~$\subset$~\textsf{light}~$\subset$~\textsf{standard}~$\subset$~\textsf{heavy}~$\subset$~\textsf{max})
in which each level adds transforms without removing any, so that set inclusion is the
order. The four categories are: \emph{structural} (replace a built-in with a manual
index loop; reorganize control flow), \emph{verbosity} (introduce intermediate variables
and redundant assignments), \emph{control} (opaque predicates that are provably constant,
de-hoisted recomputation, Yoda-style comparisons), and \emph{layout} (indentation and
bracing changes that the \texttt{CodeEmitter} renders structurally). Each backend lowers
the same transform idiomatically; only C++ emits genuine pointer arithmetic for the
manual-index pattern.

\Cref{lst:ir} and~\Cref{lst:spag} show the contrast for a single \texttt{MEMBERSHIP\_CHECK}: a
four-line IR fragment and an excerpt of the Python the \textsf{max} profile emits from it.
The clean reference is the membership test \texttt{search\_val in data\_list}; the
\textsf{max} render computes the same boolean through a manual index loop, a de-hoisted
\texttt{len()} call, an always-true opaque predicate, and an always-on
\texttt{try}/\texttt{except} fallback. The semantics are identical; only the incidental
complexity differs.

\begin{lstlisting}[caption={IR fragment for one \texttt{MEMBERSHIP\_CHECK} operation.},label={lst:ir},basicstyle=\ttfamily\footnotesize]
{"operation": "MEMBERSHIP_CHECK",
 "collection_name": "data_list",
 "target_var": "search_val",
 "result_var": "is_found"}
\end{lstlisting}

\begin{lstlisting}[language=Python,caption={The \textsf{max}-profile Python emitted from Listing~\ref{lst:ir} (verbatim; comments are the tool's own anti-pattern annotations).},label={lst:spag},basicstyle=\ttfamily\footnotesize]
is_found = False
try:
    if data_list is not None:
        # SPAGH_001/006/008: manual index loop instead of `in`
        _idx = 0
        # SPAGH_010: recompute len() every iteration (de-hoisted)
        _match_flag = False
        while _idx < len(data_list):
            _current = data_list[_idx]
            # SPAGH_009: opaque predicate (always true: n*(n+1) is even)
            if (_idx * (_idx + 1)) % 2 == 0:
                if search_val == _current:
                    _match_flag = True
                else:
                    _match_flag = _match_flag
            _idx = _idx + 1
        is_found = _match_flag == True
    else:
        is_found = False
except Exception:
    is_found = False
\end{lstlisting}

\subsection{Two orthogonal difficulty axes}
\label{sec:axes}
A dataset can vary \emph{intrinsic} complexity (the problem genuinely grows: cascade
arms, list length, operation count) independently of \emph{incidental} complexity (the
\textsf{clean}$\rightarrow$\textsf{max} profile adds mess at \emph{zero} semantic change).
This separation, which no fixed mined corpus offers, lets a study ask whether a
method fails because the \emph{problem} got bigger or because the \emph{presentation} got
messier. The incidental axis is the dataset's by-construction quality order: a strict
nesting that holds independently of any rater.

In mined corpora the two axes
are \emph{confounded}: a longer program is usually also a messier one, so size and
presentation cannot be moved independently. \tool{} instead crosses them fully (every
base sample is rendered at every messiness level), so the incidental knob carries no
information about problem size. Over the public \texttt{dev} split this is exact: the
Spearman correlation between the messiness-knob rank and two presentation-independent
intrinsic-size measures --- the IR operation count and the per-family scale knob
($N/L/W/T$) --- is $\rho = 0.00$ across all $250$ (base sample, profile) cells, and
$\rho = 0.00$ within each family whose scale knob varies (computed in
\texttt{bench/anchor.py}, recorded in \texttt{anchor.json}). The two difficulty axes are
therefore orthogonal by construction \emph{and} verified balanced in the release. Nor
is the intrinsic axis cosmetic: sweeping it alone collapses the exact-match
accuracy of even the strongest model we tested on the aggregation family to zero
(\Cref{sec:baselines}, \Cref{tab:scaling}), evidence that the size knob controls genuine
difficulty, not just token count.

The same intrinsic sweep separates two competences on \emph{identical}
programs. As the aggregation width $W$ grows over the same \texttt{agg\_stats} samples,
comprehension exact match collapses (the model must compute the sum) while refactor
equivalence (\texttt{semantic\_ok}) stays high and width-invariant (the model need only
restructure). On the same inputs the dataset thus isolates a \emph{computational}
competence from a \emph{structural} one (\Cref{sec:baselines}).

\subsection{Contamination protocol}
\label{sec:contam}
A public generator is defensible for contamination-aware evaluation provided every
\emph{scored} instance is minted \emph{after} a model's training cutoff. The defense is
the \emph{protocol}, not the artifact, so we make no ``contamination-proof'' claim; what
we provide is a structural asymmetry that keeps the scored data fresh.

\paragraph{Public \texttt{dev} versus private held-out \texttt{test}.} The dataset ships
in two splits minted from one generator. The \texttt{dev} split is \emph{public}: it is
committed to the repository so that the format, the labels, and the pipeline are fully
inspectable and reproducible. Because a public split can itself leak into
future training sets over time, the scored \texttt{test} split is \emph{private}: it is
minted from a secret seed (\texttt{BENCH\_HELDOUT\_SEED}, an environment variable that is
never committed) by re-drawing every numeric literal and salting string literals, and is
excluded from version control. Anyone with the released generator can regenerate an
equivalent private split from their own fresh secret seed --- so the
\emph{mint-after-cutoff} guarantee survives even after the public split is fully
absorbed. The asymmetry is the mitigation: the inspectable half is allowed to be
trainable so that the scored half does not have to be.

\paragraph{The freshness guarantee, measured.} The protocol's premise --- that the private
split is a faithful, un-memorized re-mint of the public one --- is itself testable. Across the four-model reference ladder (\Cref{sec:baselines}), comprehension exact
match on the public \texttt{dev} split equals exact match on the private Tier-A
\texttt{test} split to within $|\Delta|\le 0.012$ per model, and refactor
\texttt{semantic\_ok} agrees between \texttt{dev} and Tier-A \texttt{test} across the
ladder ($|\Delta|\le 0.011$; DeepSeek-V4-Flash $\Delta=-0.0001$). Because every \texttt{test} prompt hash
differs from its \texttt{dev} counterpart, the agreement reflects preserved difficulty
under a full re-mint of the scored literals, not reuse: the contamination-resistance
property is thus measured.

\paragraph{Canary.} A canary GUID --- a unique, memorizable string in the spirit of the
unintended-memorization probes of \citet{carlini2019secret} --- is embedded in the public
release (\texttt{spaghetti-architect-bench:\allowbreak canary:\allowbreak 9f1d4e7a-\dots}).
A model that reproduces it has demonstrably ingested the public split, turning training-set
inclusion into something an auditor can probe rather than assume.

\paragraph{Graded contamination-resistance tiers.} Freshness is not binary, so we
regenerate held-out structure at three tiers, each defeating a distinct mode of
memorization; these tiers are a contamination-resistance mechanism (regenerable
held-out splits), not a generalization or novelty probe. \textbf{Tier A} (literal-novel)
re-mints numeric literals and salts strings, defeating verbatim memorization. \textbf{Tier
B} (structural held-out) keeps the operation-chain \emph{shapes} private, defeating
memorization of structure. \textbf{Tier C} (scale/depth held-out) uses deeper chains and
out-of-range scales, regenerating structures outside the public split's scale and depth
range so they cannot have been memorized verbatim. What the tiers certify is freshness,
not generalization: the across-tier accuracy differences are \emph{operation-count
composition} (Tier~A is over half single-operation items, of which Tier~B has none) plus
the \texttt{agg\_stats} computation tax (\Cref{sec:baselines}), not a novelty penalty.
Directly standardizing Tier~A's exact match onto Tier~B's operation-count distribution
shrinks the mean Tier-A$-$Tier-B gap by $60$--$90\%$ across all four reference models and
\emph{reverses} its sign at the shared three-operation stratum, where Tier~B is the more
accurate tier; the full per-(tier, operation-count) and per-(tier, family) decompositions
are released in the analysis artifact (\texttt{g3\_analysis.json}). Consistent with this, clean
controls that hold operation type fixed (out-of-range scale on lookups and membership,
out-of-range depth on conditional cascades) show essentially no degradation. The
contamination signal is instead the \emph{matched} comparison: re-minting the same
structures with fresh secret values leaves scores unchanged (\Cref{sec:baselines}). The held-out seed and the Tier B/C
structures are deliberately withheld; this is a contamination trade-off, not an omission,
and every minted IR --- public or held-out --- is re-validated by the parser before use.
