%% sections/validity.tex
%% Construct validity. Numbers below are REAL, regenerated from bench/out/anchor.json
%% via `~/.venvs/spaghetti-metrics/bin/python bench/anchor.py`. Every rho is the mean over
%% the 50 dev base samples (n=50) with a nonparametric bootstrap 95% CI (grade.ci95_bootstrap,
%% 2000 resamples; key incidental_knob_spearman_ci95). The convergent-metric table
%% (Table~\ref{tab:validity}) is computed on the PYTHON sources; lizard is the one
%% language-agnostic anchor and is validated per language across all five on TWO facets ---
%% cyclomatic complexity AND lexical token count (anchor.json cross_language.per_language /
%% per_language_tokens; Table~\ref{tab:validity-xlang}). A missing tool records an honest
%% SKIP, never a fabricated number. radon CC/MI, cognitive complexity and the Buse--Weimer
%% feature re-implementation are Python-only tools and are reported as such; the non-circular
%% readability leg (BW +0.39) stays Python-only by design (the human study would extend it).

\section{Construct Validity: do the labels mean anything?}
\label{sec:validity}

The by-construction quality order is only useful if it tracks quantities the community
already trusts. We test whether the incidental knob moves established static metrics
monotonically, over the public \texttt{dev} set ($50$ base samples expanded to $300$
sources across the six knob levels). For each base sample we hold semantics fixed and
compute Spearman's $\rho$ between the knob rank
(\texttt{clean}${}\prec{}$\texttt{minimal}${}\prec{}$\texttt{light}${}\prec{}$%
\texttt{standard}${}\prec{}$\texttt{heavy}${}\prec{}$\texttt{max}) and each metric,
reporting the mean over samples. The anchor tools
(\texttt{radon}~6.0.1, \texttt{lizard}~1.23.0, \texttt{cognitive\_complexity}~1.3.0)
are kept off the zero-dependency core and run only inside a quarantined virtualenv; a
missing tool records an honest \texttt{SKIP} rather than a fabricated number.

We are explicit about what is measured on which language, because our headline is a
five-language parallel corpus. Three of the four anchors are Python-only tools:
\texttt{radon} cyclomatic complexity and maintainability index, and
\texttt{cognitive\_complexity}, all operate on the Python abstract syntax tree, and our
Buse--Weimer feature re-implementation tokenises Python. Table~\ref{tab:validity} therefore
reports the convergent-metric evidence \emph{on the Python sources}. Only \texttt{lizard}
cyclomatic complexity is genuinely language-agnostic, so we additionally run it
\emph{per language} on all five sources (Table~\ref{tab:validity-xlang}) to check that the
complexity signal is reproduced in every language and is not an artifact of measuring
Python.

\begin{table}[t]
\centering
\caption{Spearman correlation between the incidental ``messiness'' knob rank and
established static metrics \emph{on the Python sources}, as the mean over the $50$
\texttt{dev} base samples ($n{=}50$) with a nonparametric bootstrap $95\%$ CI ($2000$
resamples over base samples; higher $|\rho|$ = the label order is better reflected).
Complexity metrics move strongly with the knob, but partly \emph{by construction}; the
less tautological readability signal is markedly weaker, and its CI sits clearly below
the complexity metrics' while still excluding zero. \texttt{radon}, cognitive complexity
and the Buse--Weimer features are Python-only tools; the cross-language check for
\texttt{lizard} is in Table~\ref{tab:validity-xlang}.}
\label{tab:validity}
\begin{tabular}{@{}llr@{}}
\toprule
Metric (tool) & Direction & Spearman $\rho$ vs.\ knob ($95\%$ CI) \\
\midrule
Cyclomatic complexity (radon, Python)   & + & $+0.75\ [+0.72,+0.78]$ \\
Cyclomatic complexity (lizard, Python)  & + & $+0.75\ [+0.72,+0.78]$ \\
Cognitive complexity (Python)           & + & $+0.80\ [+0.76,+0.84]$ \\
Maintainability index (radon, Python)   & $-$ & $-0.79\ [-0.86,-0.70]$ \\
Buse--Weimer surface density (Python, agg.) & + & $+0.39\ [+0.29,+0.49]$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Complexity moves strongly, but read this with care.}
On the Python sources the complexity anchors agree and move sharply with the knob:
cyclomatic complexity~\citep{mccabe1976complexity} (identical at $\rho=+0.75$ under both
\texttt{radon} and \texttt{lizard}), cognitive complexity~\citep{campbell2018cognitive}
($\rho=+0.80$), and the maintainability index in
the expected \emph{opposite} direction ($\rho=-0.79$). Two independent tool families
converging on the same ordering --- with tight bootstrap CIs that do not overlap the
readability signal's (Table~\ref{tab:validity}) --- and our own stdlib metric lane
cross-checking against them at $\rho>0.97$, is reassuring evidence that the labels are not
arbitrary. We
deliberately under-claim here, though, because this agreement is partly
\emph{tautological}: the incidental knob is engineered to inject redundant control flow,
dead branches, and nested scaffolding, which is precisely what cyclomatic and cognitive
complexity count. A metric rising when we add the very constructs it was designed to
detect confirms that the transformation does what it says, but it is weaker evidence that
the order matches \emph{human} judgments of quality.

\paragraph{Two convergent signals reproduce in all five languages.}
Because three of the four anchors are Python-only, Table~\ref{tab:validity} could in
principle be a Python-specific accident. We rule that out with \texttt{lizard}, which
lexes each language natively, and report \emph{two} of its facets per language over the
five engine ``messiness'' profiles
(\texttt{minimal}${}\prec{}$\texttt{light}${}\prec{}$\texttt{standard}${}\prec{}$%
\texttt{heavy}${}\prec{}$\texttt{max}; the \texttt{clean} idiomatic floor is Python-only,
so it is excluded here and reported separately). First, cyclomatic complexity rises with
the knob in \emph{every} language ($\rho\approx+0.34$ to $+0.35$, CIs clear of zero, no
\texttt{SKIP}); this is the weaker, partly-circular facet --- modest within the spaghetti
band because branch-bearing control flow is injected mainly at \texttt{max}. Second, and
more strongly, \texttt{lizard}'s own lexical token count --- a genuine per-language size
measure --- rises with the knob at $\rho=+0.88$ $[+0.85,+0.92]$ in all five languages
(Table~\ref{tab:validity-xlang}). The values are near-identical across languages because
the five sources are rendered from \emph{one} IR (a parallel corpus by construction;
\texttt{go}'s slightly lower CC is the only divergence). We are careful about what this
does and does not show: it is strong evidence that the incidental transformation
\emph{manifests faithfully in every language} --- the messiness is genuinely present,
ordered, and measurable cross-language, not a Python artifact --- but token size and
cyclomatic complexity are surface/structural facets, \emph{not} human quality. The
genuinely less-circular readability signal remains Python-only (next paragraph), and that
$\rho=+0.39$, not these cross-language faithfulness signals, is our conservative quality
bound.

\begin{table}[t]
\centering
\caption{Cross-language construct validity: Spearman correlation between the incidental
``messiness'' knob rank and two \texttt{lizard} facets computed \emph{per language} ---
cyclomatic complexity and lexical token count --- over the five engine profiles
(\texttt{minimal}${}\prec{}$\texttt{light}${}\prec{}$\texttt{standard}${}\prec{}$%
\texttt{heavy}${}\prec{}$\texttt{max}), as the mean over the $50$ \texttt{dev} base samples
with a bootstrap $95\%$ CI. \texttt{lizard} lexes each language natively. Both facets are
positive in all five languages; token size is the stronger, tighter signal. Python's
$+0.75$ CC in Table~\ref{tab:validity} additionally includes the \texttt{clean} idiomatic
baseline, only definable for Python (its within-profile gradient is the $+0.35$ here).}
\label{tab:validity-xlang}
\begin{tabular}{@{}lrr@{}}
\toprule
Language & lizard CC $\rho$ ($95\%$ CI) & lizard token-size $\rho$ ($95\%$ CI) \\
\midrule
Python      & $+0.35\ [+0.26,+0.45]$ & $+0.88\ [+0.85,+0.92]$ \\
JavaScript  & $+0.35\ [+0.26,+0.45]$ & $+0.88\ [+0.85,+0.92]$ \\
Go          & $+0.34\ [+0.24,+0.44]$ & $+0.88\ [+0.85,+0.92]$ \\
Java        & $+0.35\ [+0.26,+0.45]$ & $+0.88\ [+0.85,+0.92]$ \\
C++         & $+0.35\ [+0.26,+0.45]$ & $+0.88\ [+0.85,+0.92]$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{The readability signal is real but weaker.}
The more informative test is the one anchor that is \emph{not} a complexity counter. Our
re-implementation of the Buse--Weimer~\citep{buse2010readability} surface readability
feature set (computed with the Python stdlib over the Python sources) correlates with the
knob at only $\rho=+0.39$ $[+0.29,+0.49]$ in the documented density aggregate --- roughly
half the strength of the complexity metrics, with a bootstrap CI clearly below theirs. We
read this honestly: the knob makes code denser and
harder to read, but far less unambiguously than it inflates control-flow complexity, and
even this number is a surface-density proxy rather than the original trained readability
classifier (whose coefficients are not available offline, so that anchor remains a
recorded \texttt{SKIP}). We compute these surface features on Python only, matching the
other density anchors; we do not extrapolate a readability number to the other four
languages.

\paragraph{No human calibration yet: a deliberate scope boundary, and future work.}
We are explicit about the limit of all of the above: the benchmark's construct validity
rests entirely on convergent \emph{static} metrics, not on calibrated human judgment. We
make no claim that the by-construction labels have been validated against human
perception of quality or readability. The complexity anchors are partly tautological, and
the single least-tautological static signal (the Buse--Weimer surface-density proxy,
$\rho=+0.39$) is a re-implemented feature set, not a human-rated model. We therefore treat
the $\rho=+0.39$ readability signal, not the $\rho=+0.80$ complexity signal, as the
conservative bound on how much the by-construction order reflects perceived quality. A
human-rating study calibrated to this exact stimulus set is the natural next step and is
left as explicit future work; concretely, we plan to recruit $N\approx15$--$20$
experienced developers to blind-rank a stratified subset of (sample, knob) pairs ---
balanced across families, intrinsic scales, and the five languages --- and to measure the
rank agreement (Kendall's $W$ and mean per-rater Spearman $\rho$) between the human
orderings and the by-construction knob order, pre-registered and reported per language.
Until that study is run, we do not assert that the labels track human judgment; we assert
only that they track convergent, independently-implemented static metrics --- structural
complexity and lexical size --- in every language, and a less-circular surface-readability
signal in Python. Extending a \emph{non-circular} signal to the other four languages is
precisely what the human study would add, and we scope our claims to that boundary.
