%% Spaghetti Architect — DMLR (Journal of Data-centric Machine Learning Research),
%% data.mlr.press: the rolling JMLR-family archival journal (NOT the dmlr.ai workshops).
%%
%% SUBMISSION REQUIREMENTS — confirmed 2026-06-26 from data.mlr.press/submissions.html:
%%   - NO page limit (longer papers simply take longer to review); we are ~16pp.
%%   - SINGLE-BLIND + open review: authors are NOT anonymized (name/affiliation shown).
%%   - MUST use this official dmlr2e style (failure to adhere may be desk-rejected).
%%   - Submit via OpenReview: https://openreview.net/group?id=DMLR
%%   - A Broader Impact Statement is REQUIRED  -> see \impact{} below.
%%   - Funding + competing-interests disclosure is REQUIRED -> see \acks{} below.
%%   - Dataset papers: supplementary must include a datasheet (Appendix + DATASHEET.md),
%%     a reviewer-accessible download URL (the public GitHub repo), an author-liability +
%%     data-license statement (datasheet "Author statement"), and a hosting/licensing/
%%     maintenance plan (datasheet "Maintenance").
%%
%% STILL TODO (maintainer action): author email + affiliation; Zenodo archival DOI; the
%%   funding/competing-interests text in \acks; finalize \dmlrheading once a paper id is
%%   assigned at acceptance.
%%
%% BUILD:  pdflatex dmlr -> bibtex dmlr -> pdflatex dmlr -> pdflatex dmlr
%%   (requires dmlr2e.sty in this directory — vendored from
%%    github.com/JmlrOrg/dmlr-style-file; sample.tex/sample.bib are the upstream sample.)
%%
%% SCOPE: this is the DATA-CENTRIC RESOURCE paper — the generator + the contamination-
%% resistant, by-construction-labelled, multi-language code dataset + label-validity
%% evidence + reference baselines. It MUST NOT report the LLM-as-judge meta-evaluation
%% FINDINGS: that empirical study is a SEPARATE, deferred study (seed draft archived at
%% archived/benchmark-paper/). Cite it only as a motivating use case; never put its
%% (unrun) numbers here.

\documentclass[twoside,11pt]{article}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}              % load before hyperref (which dmlr2e pulls in)

% Official DMLR style — vendored from github.com/JmlrOrg/dmlr-style-file.
% Provides epsfig, amssymb, natbib, graphicx, hyperref; sets bib style to plainnat.
% Any additional packages must be loaded AFTER it.
\usepackage{dmlr2e}

\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{tikz}
\usetikzlibrary{positioning,fit,backgrounds,shapes.geometric}
\usepackage{listings}
\usepackage{cleveref}             % after hyperref; names references ("Section", ...) itself
\crefname{lstlisting}{Listing}{Listings}
\Crefname{lstlisting}{Listing}{Listings}
\crefname{appendix}{Appendix}{Appendices}
\Crefname{appendix}{Appendix}{Appendices}

\lstset{basicstyle=\ttfamily\small,breaklines=true,frame=single,columns=fullflexible}

% The datasheet appendix is dense with \texttt identifiers/paths; let TeX stretch a line
% as a last resort (only when it cannot otherwise be set) to absorb small overfulls in
% the narrower DMLR column without loosening normal paragraphs.
\setlength{\emergencystretch}{3em}

% editorial markup — STRIP before camera-ready (a single \renewcommand to {} hides them)
\newcommand{\todo}[1]{\textcolor{red}{\textbf{[TODO: #1]}}}
\newcommand{\owner}[1]{\textcolor{blue}{\textsf{[\textbf{owner:} #1]}}}
\newcommand{\port}[1]{\textcolor{orange}{\textsf{[port: #1]}}}

\newcommand{\tool}{\textsc{Spaghetti Architect}}

% DMLR running-head metadata — placeholders until acceptance/camera-ready.
% args: {volume}{year}{pages}{submitted}{published}{paper id}{author full names}
% (volume/pages/id are assigned at acceptance; lastpage.sty is absent here, so pages are
%  left blank rather than auto-computed.)
\dmlrheading{}{2026}{}{6/2026}{}{}{Yuxiang Ji}
\ShortHeadings{Spaghetti Architect}{Ji}
\firstpageno{1}
\def\openreview{\url{https://openreview.net/forum?id=XXXX}} % camera-ready: real OpenReview forum id

\begin{document}

\title{\tool: A Contamination-Resistant, By-Construction-Labelled,\\
       Multi-Language Code Dataset Generator}

\author{\name Yuxiang Ji \email kurathandrew@gmail.com\\
       \addr Sunway College}

\editor{}

\maketitle

\begin{abstract}%
Mining-based code corpora are abundant but \emph{uncontrolled}: a snippet's semantics,
its surface ``messiness,'' and its difficulty are whatever the wild happened to
contain, there is no known-optimal reference to grade against, and any public sample
may already sit in a model's training set. We present \tool, a tool that mints code
datasets with exactly the control such corpora lack. An \emph{anti-optimization
transpiler} maps a clean, language-agnostic JSON intermediate representation to
deliberately redundant, fully-flattened programs in five languages
(Python, JavaScript, Go, Java, C++); every emitted program is compiled, run, and
checked against a reference oracle, so each instance is correct \emph{by construction}.
Because the clean IR is a known-optimal reference and ``messiness'' is dialed by
composable, strictly-nested anti-pattern profiles, the tool labels each instance along
two \emph{orthogonal} difficulty axes --- intrinsic (problem size) and incidental
(presentation at fixed semantics) --- and resists contamination by minting fresh
variants from a private held-out seed. We document the generator and the dataset it
produces, give construct-validity evidence that the by-construction quality order moves
established complexity and readability metrics, and report reference baselines on a
four-model open capability ladder --- where exact-match accuracy tracks model capability,
and the intrinsic scale knob collapses even the frontier model's arithmetic-aggregation
accuracy to zero. Two further results bear out the design. First, the held-out protocol
resists contamination not only by construction: development-set scores equal a
freshly re-minted private test split across the ladder to within $|\Delta|\le 0.011$, for
both comprehension (exact match) and refactoring (semantic-equivalence rate). Second, on
identical programs the dataset separates two competences: refactoring
equivalence (itself a capability ladder, $0.73\!\rightarrow\!0.99$) is invariant to the
intrinsic scale knob for the capable models while output-prediction
accuracy collapses with it. One instance set therefore probes structural and computational
ability independently. The artifact is open source (MIT), dependency-free, and will be archived
under a DOI.
\end{abstract}

\begin{keywords}
  data-centric machine learning, code dataset generation, data contamination,
  by-construction labels, multi-language code, benchmark resource, dataset documentation
\end{keywords}

\input{sections/intro}
\input{sections/resource}
\input{sections/validity}
\input{sections/baselines}
\input{sections/related}
\input{sections/limitations}

\impact{%
\tool{} is research infrastructure for \emph{evaluation}, not a model or a deployed
system, and its intended impact is methodological: contamination-resistant,
by-construction-labelled data lets the community measure code-model behavior (code
quality, robustness to incidental complexity, and output comprehension) with controls
that mined corpora cannot supply, while the held-out, regenerable design reduces the
benchmark leakage that inflates reported progress. The most direct risk is dual use: the
generator deliberately emits degraded-but-correct programs, which could in principle be
repurposed to obfuscate software or to manufacture low-quality training data. We judge
this risk low: the outputs are intentionally verbose, oracle-checked transformations
of a public IR rather than novel capabilities. We nonetheless release the
tool under a permissive license with an explicit research framing, and document the
canary and held-out protocol so that misuse for contamination is detectable rather than
silent. The generated data are wholly synthetic and contain no personal, sensitive, or
human-subject content.
}

\acks{%
This work received no specific grant or external funding. The author declares no
competing interests.
}

\bibliography{refs}

\appendix
\crefalias{section}{appendix}
\input{sections/datasheet}

\end{document}
